{\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of finite-dimensional and affine Lie algebras

### 2011-07-15 doc/paper.tex
Examples in paper completed.

\newcommand{\go}{\stackrel{\circ }{\mathfrak{g}}}
\newcommand{\ao}{\stackrel{\circ }{\mathfrak{a}}}
\newcommand{\co}[1]{\stackrel{\circ }{#1}}
\newcommand{\pia}{\pi_{\mathfrak{a}}}
\newcommand{\piab}{\pi_{\mathfrak{a}_{\bot}}}
\newcommand{\gf}{\mathfrak{g}}
\newcommand{\af}{\mathfrak{a}}
\newcommand{\bff}{\mathfrak{b}}
\newcommand{\afb}{\mathfrak{a}_{\bot}}
\newcommand{\hf}{\mathfrak{h}}
\newcommand{\hfg}{\hf_{\gf}}
\newcommand{\hfb}{\mathfrak{h}_{\bot}}
\newcommand{\pf}{\mathfrak{p}}

@@

The structure of affine Lie algebras allows to extend computational algorithms created for finite-dimensional Lie algebras  \cite{Fuchs:1996dd}, \cite{gannon2001algorithms}, \cite{kass1990ala}. The book \cite{kass1990ala} with the tables of multiplicities and other computed characteristics of affine Lie algebras and representations was published in 1990. But we are not aware of software packages for popular computer algebra systems which can be used to extend these results.
We address this issue and present {\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of affine and finite-dimensional Lie algebras.  We describe the features and limitations of the package in present paper and in the manual \cite{affinemanual}. In present paper we also provide representation-theoretical background of implemented algorithms and present some examples of computations relevant to physics.

@@

Affine weight is an extension of a finite weight by supplying it
with level and grade coordinates. There is a set of functions
defined for finite and affine weights. The complete list can be
found in the user manual \cite{affinemanual}. The most important
are the definitions of addition, multiplication by number and
scalar product (bilinear form) for weights. These definitions
allow to use traditional notation when working with {\bf
Affine.m}:
\begin{lstlisting}
  w=makeFiniteWeight[{1,0,3}];
  v=makeFiniteWeight[{3,2,1}];
  2*w+v==makeFiniteWeight[{5,2,7}]
  w.v==6
\end{lstlisting}

@@

{\bf Affine.m} can be used to study different kinds of modules, i.e. Verma modules, irreducible modules and parabolic Verma modules .  We need datastructure \lstinline{module} to represent generic module of Lie algebra $\gf$. Module properties can be deduced from its set of singular weights using Weyl character formulae \eqref{eq:11},\eqref{eq:12},\eqref{eq:18},\eqref{eq:13}. Set of singular weights can have Weyl symmetry. It can be symmetry with respect to Weyl group $W_{\gf}$ or with respect to some subalgebra $W_{\af}$ as in the case of parabolic Verma modules. Then it is possible to study only main Weyl chamber $C_{\af}$. To use this symmetry generic constructor for \lstinline{module} datastructure accepts several parameters \lstinline{makeModule[rs_?rootSystemQ][singWeights_formalElement,subs_?rootSystemQ|emptyRootSystem[],opts___?OptionQ]}. Here \lstinline{rs} is root system of Lie algebra $\gf$,\lstinline{singWeights} is the set of singular weights,  \lstinline{subs} is root system corresponding to Weyl group $W_{\af}$ which is the symmetry of the set of singular weights.
There are several specialized constructors for different types of highest weight modules:
\begin{lstlisting}[mathescape=true]
vm=makeVermaModule[$B_{2}$][weight[$B_{2}$][2,1]];
pm=makeParabolicVermaModule[$B_{2}$][weight[$B_{2}$][2,1],{1}];
im=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][2,1]];
GraphicsRow[textPlot/@{im,vm,pm}]
$\includegraphics[width=130mm]{irrep-verma-pverma.pdf}$
\end{lstlisting}

As we already stated properties of the module are encoded by its singular element. Function \lstinline{singularElement[m_module]} returns singular element of module as \lstinline{formalElement} datastructure. Character (up to \lstinline{limit} for (parabolic) Verma modules) is returned by function \lstinline{character[m_module]}. Direct sum of modules is module and we use natural notation
\begin{lstlisting}[mathescape=true]
im1=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][2,1]];
im2=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][1,2]];
textPlot[im1$\oplus$ im2]
$\includegraphics[width=60mm]{irrep-sum.pdf}$
\end{lstlisting}

@@

\begin{figure}[h]
  \noindent\centering{
    \includegraphics[width=80mm]{timing.pdf}
  }
  \caption{Running time of algorithms based on recurrent formula \eqref{eq:15} (in blue) and \eqref{eq:14} (in red) with the number of weights in $\bar C$ for calculation of multiplicities in representations of $B_{2}$.}
\label{fig:freudenthal-racah-times}

@@

\begin{figure}[h]
  \noindent\centering{
   \includegraphics[width=100mm]{branching-timing.pdf}
  }
  \caption{Running time of algorithms based on recurrent formula \eqref{eq:15} (in blue) and \eqref{eq:14} (in red) with the number of weights in $\bar C$ for calculation of branching coefficients for $B_{3}\subset B_{4}$.}
  \label{fig:branching}
\end{figure}

@@

We illustrate the connection of branching with generalized BGG-resolution with the sequence of parabolic Verma modules which appear in the study of branching of irreducible $B_{2}$-modules to the modules of subalgebra $A_{1}$. Root of the subalgebra $A_{1}$ in this case is equal to the highest root $\Theta$ of algebra $B_{2}$. In this case orthogonal sub-algebra $\afb$ gives rise to the parabolic Verma modules \cite{2011arXiv1102.1702L}.

\begin{figure}[h]
  \label{branching-bgg}
  \noindent\centering{
    \includegraphics[width=80mm]{branching.pdf}
  }
  \caption{Parabolic Verma modules from branching}
\end{figure}

@@

Branching functions of the embedding $\af\subset\gf$ are the partition functions of CFT on the torus.

\begin{lstlisting}[mathescape=true]
branchingFunctions[$\hat B_{2}$,makeAffineExtension[makeFiniteRootSystem[{{1, 1}}]], {1, 1, 1}]

@@

\begin{figure}[h]
  \label{branching-b4-b2}
  \noindent\centering{
    \includegraphics[width=80mm]{branching.pdf}
  }
  \caption{Branching coefficients of module $L^{\left[0,1,0,2\right]}$ for regular embedding $B_{2}\subset B_{4}$ displayed in the basis $\left\{e_3,e_4\right\}$}
\end{figure}

\section{Conclusion}
\label{sec:conclusion}
We present the package {\bf Affine.m} for computations in representation theory of finite-dimensional and affine Lie algebras. The package is available from {\it github} and is also submitted to the library.

@@

\section{Software package}
\label{package}
The package can be freely downloaded from {\it github}. Latest release can be found here. To get the development code use the command
\begin{lstlisting}[language=bash]
 git clone git://github.com/naa/Affine.git
\end{lstlisting}

Contents of the package:
\begin{verbatim}
    Affine/                                root folder
      demo/                                  demonstrations
        demo.nb                                demo notebook
        paper.nb                               code from the paper
      doc/                                 documentation folder
        figures/                             figures in paper and manual
          timing.pdf                           diagram showing performance
          branching.pdf                        branching coefficients
        bibliography.bib                     bibliographic database
        manual.pdf                           manual
        manual.tex                           manual source
        paper.pdf                            paper
        paper.tex                            paper source
        TODO.org                             list of issues
      src/                                 source folder
        affine.m                             main software package
      tests/                               unit tests folder
        tests.m                              unit tests
      README.markdown                      short description of software
\end{verbatim}

>>>>

\newcommand{\go}{\stackrel{\circ }{\mathfrak{g}}}
\newcommand{\ao}{\stackrel{\circ }{\mathfrak{a}}}
\newcommand{\co}[1]{\stackrel{\circ }{#1}}
\newcommand{\pia}{\pi_{\mathfrak{a}}}
\newcommand{\piab}{\pi_{\mathfrak{a}_{\bot}}}
\newcommand{\gf}{\mathfrak{g}}
\newcommand{\af}{\mathfrak{a}}
\newcommand{\bff}{\mathfrak{b}}
\newcommand{\afb}{\mathfrak{a}_{\bot}}
\newcommand{\hf}{\mathfrak{h}}
\newcommand{\hfg}{\hf_{\gf}}
\newcommand{\hfb}{\mathfrak{h}_{\bot}}
\newcommand{\pf}{\mathfrak{p}}
\newcommand{\aft}{\widetilde{\mathfrak{a}}}

@@

The structure of affine Lie algebras allows to extend computational algorithms created for finite-dimensional Lie algebras  \cite{Fuchs:1996dd}, \cite{gannon2001algorithms}, \cite{kass1990ala}. The book \cite{kass1990ala} with the tables of multiplicities and other computed characteristics of affine Lie algebras and representations was published in 1990. But we are not aware of software packages for popular computer algebra systems which can be used to extend these results.
We address this issue and present {\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of affine and finite-dimensional Lie algebras.  We describe the features and limitations of the package in present paper.  We also provide representation-theoretical background of implemented algorithms and present some examples of computations relevant to physics.

@@

Affine weight is an extension of a finite weight by supplying it
with level and grade coordinates. There is a set of functions
defined for finite and affine weights. The complete list can be
found in package online help. The most important
are the definitions of addition, multiplication by number and
scalar product (bilinear form) for weights. These definitions
allow to use traditional notation when working with {\bf
Affine.m}:
\begin{lstlisting}
  w=makeFiniteWeight[{1,0,3}];
  v=makeFiniteWeight[{3,2,1}];
  2*w+v==makeFiniteWeight[{5,2,7}]
  w.v==6
\end{lstlisting}

@@

{\bf Affine.m} can be used to study different kinds of modules, i.e. Verma modules, irreducible modules and parabolic Verma modules .  We need datastructure \lstinline{module} to represent generic module of Lie algebra $\gf$. Module properties can be deduced from its set of singular weights using Weyl character formulae \eqref{eq:11},\eqref{eq:12},\eqref{eq:18},\eqref{eq:13}. Set of singular weights can have Weyl symmetry. It can be symmetry with respect to Weyl group $W_{\gf}$ or with respect to some subalgebra $W_{\af}$ as in the case of parabolic Verma modules. Then it is possible to study only main Weyl chamber $C_{\af}$. To use this symmetry generic constructor for \lstinline{module} datastructure accepts several parameters \lstinline{makeModule[rs_?rootSystemQ][singWeights_formalElement,subs_?rootSystemQ|emptyRootSystem[],opts___?OptionQ]}. Here \lstinline{rs} is root system of Lie algebra $\gf$,\lstinline{singWeights} is the set of singular weights,  \lstinline{subs} is root system corresponding to Weyl group $W_{\af}$ which is the symmetry of the set of singular weights.
There are several specialized constructors for different types of highest weight modules:
\begin{lstlisting}[mathescape=true]
vm=makeVermaModule[$B_{2}$][weight[$B_{2}$][2,1]];
pm=makeParabolicVermaModule[$B_{2}$][weight[$B_{2}$][2,1],{1}];
im=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][2,1]];
GraphicsRow[textPlot/@{im,vm,pm}]
$\includegraphics[width=130mm]{figures/irrep-verma-pverma.pdf}$
\end{lstlisting}

As we already stated properties of the module are encoded by its singular element. Function \lstinline{singularElement[m_module]} returns singular element of module as \lstinline{formalElement} datastructure. Character (up to \lstinline{limit} for (parabolic) Verma modules) is returned by function \lstinline{character[m_module]}. Direct sum of modules is module and we use natural notation
\begin{lstlisting}[mathescape=true]
im1=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][2,1]];
im2=makeIrreducibleModule[$B_{2}$][weight[$B_{2}$][1,2]];
textPlot[im1$\oplus$ im2]
$\includegraphics[width=60mm]{figures/irrep-sum.pdf}$
\end{lstlisting}

@@

\begin{figure}[h]
  \noindent\centering{
    \includegraphics[width=80mm]{figures/timing.pdf}
  }
  \caption{Running time of algorithms based on recurrent formula \eqref{eq:15} (in blue) and \eqref{eq:14} (in red) with the number of weights in $\bar C$ for calculation of multiplicities in representations of $B_{2}$.}
\label{fig:freudenthal-racah-times}

@@

\begin{figure}[h]
  \noindent\centering{
   \includegraphics[width=100mm]{figures/branching-timing.pdf}
  }
  \caption{Running time of algorithms based on recurrent formula \eqref{eq:15} (in blue) and \eqref{eq:14} (in red) with the number of weights in $\bar C$ for calculation of branching coefficients for $B_{3}\subset B_{4}$.}
  \label{fig:branching}
\end{figure}

@@

We illustrate generalized BGG-resolution with the diagrams of $G_{2}$ parabolic Verma modules which appear in the decomposition of irreducible module $L^{[1,1]}_{G_{2}}$:
\begin{equation}
\mathrm{ch}\left( L^{\mu }\right) =\sum_{u\in U}\;e^{\mu _{\aft}\left(
u\right) }\epsilon (u)\mathrm{ch}M_{I}^{\mu _{\frak{a}_{\perp }}\left(
u\right) }.  \label{char-in-gen-verma-mod}
\end{equation}
Character of $L^{[1,1]}$ is presented in Figure \ref{branching-bgg}, characters of generalized Verma modules in decomposition \eqref{char-in-gen-verma-mod} are shown in Figure \ref{g2-pverma}. Characters in upper row appear in \eqref{char-in-gen-verma-mod} with positive sign and in lower row with negative.

\begin{figure}[h]
  \label{branching-bgg}
  \noindent\centering{
    \includegraphics[width=80mm]{figures/G2-irrep.pdf}
  }
  \caption{Character of irreducible $G_{2}$-module $L^{[1,1]}$}
\end{figure}
\begin{figure}[h]
  \label{g2-pverma}
  \noindent\centering{
    \includegraphics[width=150mm]{figures/G2-pverma.pdf}
  }
  \caption{Character of $G_{2}$ generalized Verma modules appearing in decomposition of $L^{[1,1]}$}
\end{figure}

@@

Branching functions of the embedding $\af\subset\gf$ are the partition functions of CFT on the torus (see \cite{difrancesco1997cft}).

As first example we show computation of branching functions for the embedding $\hat A_{1}\to \hat B_{2}$ up to tenth grade:
\begin{lstlisting}[mathescape=true]
branchingFunctions[$\hat B_{2}$,makeAffineExtension[makeFiniteRootSystem[{{1, 1}}]], {1, 1, 1}]

@@

Another example is computation of branching functions for the regular embedding $\hat B_{2}\subset \hat C_{3}$:
\begin{lstlisting}[mathescape=true]
sub=makeAffineExtension[parabolicSubalgebra[$C_{3}$][2,3]];
branchingFunctions[$\hat C_{3}$,sub, {2, 0, 0, 0}]

{{0, 1, 0}, 
  $2 q - 20 q^3 + 24 q^4 + 82 q^5 - 320 q^6 + 108 q^7$}, 
{{1, 0, 0}, 
  $1 - q - 8 q^2 + 19 q^3 + 16 q^4 - 156 q^5 + 205 q^6 + 640 q^7$}, 
{{0, 0, 1}, 
  $q - 5 q^3 + 7 q^5$}
\end{lstlisting}

\section{Conclusion}
\label{sec:conclusion}
We have presented the package {\bf Affine.m} for computations in representation theory of finite-dimensional and affine Lie algebras. It can be used to study Weyl groups, roots systems, irreducible, Verma and parabolic Verma modules of finite-dimensional and affine Lie algebras.  In present paper we have also discussed main ideas used for implementation of the package and described most important notions of representation theory required to use {\bf Affine.m}.

We have demonstrated that recurrent approach based upon Weyl character formula is not only useful for calculations but also allows to see connection with (generalized) Bernstein-Bernstein-Gelfand resolution.

Also we have presented examples of computations with this package connected with problems of physics and mathematics.

\section*{Acknowledgements}
\label{sec:acknowledgements}
The work is supported by the Chebyshev Laboratory
(Department of Mathematics and Mechanics, Saint-Petersburg State
University) under the grant 11.G34.31.0026 of the Government of the
Russian Federation.

@@

\section{Software package}
\label{package}
The package can be freely downloaded from \url{http://github.com/naa/Affine}. To get the development code use the command
\begin{lstlisting}[language=bash]
 git clone git://github.com/naa/Affine.git
\end{lstlisting}

Contents of the package:
\begin{verbatim}
    Affine/                                root folder
      demo/                                  demonstrations
        demo.nb                                demo notebook
        paper.nb                               code for the paper
      doc/                                 documentation folder
        figures/                             figures in paper 
          timing.pdf                           diagram showing performance
          branching-timing.pdf                 ...  for branching coefficients  
          irrep-sum.pdf                        sum of B2 irreps
          irrep-verma-pverma.pdf               irrep, Verma, (p)Verma for B2
          G2-irrep.pdf                         irrep for G2
          G2-pverma.pdf                        parabolic Verma for G2
        bibliography.bib                     bibliographic database
        paper.pdf                            present paper
        paper.tex                            paper source
        TODO.org                             list of issues
      src/                                 source folder
        affine.m                             main software package
      tests/                               unit tests folder
        tests.m                              unit tests
      README.markdown                      installation and usage notes
\end{verbatim}
